---
title: "assignment"
output: html_document
date: "2026-04-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# 1. Load all required packages
suppressPackageStartupMessages({
  library(epiwraps)
  library(ggplot2)
  library(AnnotationHub)
  library(ensembldb)
  library(edgeR)
  library(rtracklayer)
  library(BiocParallel)
  library(bsseq)   
  library(DMRcate) 
  library(pheatmap) 
})

```
```{r}

if (!file.exists("ESC_astrocytes_BSseq.rds")) {
  
  cat("Data not found. Downloading now (this might take a minute)...\n")
  options(timeout = 6000) 
  
  
  if (!file.exists("w11_practical.zip")) {
    download.file("https://ethz-ins.org/content/w11_practical.zip", "w11_practical.zip")
  }
  
  unzip("w11_practical.zip")
  cat("Download and extraction complete!\n")
  
} else {
  cat("Data already present! Skipping download.\n")
}
```

```{r}
bs <- readRDS("ESC_astrocytes_BSseq.rds")

genes <- readRDS("hg19_chr1_genes.rds")

bs
```
```{r}

if (!file.exists("dmr_chr1.rds")) {
  cat("Calculating chr1 DMRs (this will be much faster!)...\n")
  
  bs_filtered <- bs[rowSums(assay(bs, "Cov") > 0) == ncol(bs), ]
  
  bs_filtered <- bs_filtered[seqnames(bs_filtered) == "chr1", ]
  
  design <- model.matrix(~celltype, data = pData(bs_filtered)) 
  methdesign <- modelMatrixMeth(design)
  
  bs_smoothed <- BSmooth(bs_filtered, BPPARAM = MulticoreParam(4, progress = TRUE))
 
  seqAnnot <- sequencing.annotate(bs_smoothed, methdesign, all.cov = TRUE, coef = "celltypeastrocytes")
  
  dmrcateRes <- dmrcate(seqAnnot, C = 2, min.cpgs = 5)
  dmrRanges <- extractRanges(dmrcateRes, genome = "hg19")
  
  saveRDS(dmrRanges, "dmr_chr1.rds")
  cat("chr1 DMRs calculated and saved!\n")
  
} else {
  cat("chr1 DMRs already calculated! Loading from file...\n")
  dmrRanges <- readRDS("dmr_chr1.rds")
}

head(dmrRanges)
```

#Task 1 Chunk
```{r}

top5_dmrs <- head(dmrRanges[order(abs(dmrRanges$meandiff), decreasing = TRUE)], 5)

overlapping_genes <- subsetByOverlaps(genes, top5_dmrs)

met_genes <- bsseq::getMeth(bs, regions = overlapping_genes, type = "raw", what = "perRegion")

rownames(met_genes) <- overlapping_genes$symbol

pheatmap(met_genes, 
         main = "Methylation of Genes in Top 5 chr1 DMRs",
         cluster_cols = FALSE) # Keeps ESC and Astrocyte columns grouped together
```
#Task 2
```{r}

if (!file.exists("ATAC_AstroVsESC_subset.rds")) {
  cat("Downloading ATAC-seq results...\n")
  download.file("https://www.ethz-ins.org/content/ATAC_AstroVsESC_subset.rds", 
                "ATAC_AstroVsESC_subset.rds")
} else {
  cat("ATAC-seq data already present! Skipping download.\n")
}

atac_res <- readRDS("ATAC_AstroVsESC_subset.rds")

atac_chr1 <- atac_res[seqnames(atac_res) == "chr1"]

head(atac_chr1)
```

```{r}

overlaps <- findOverlaps(dmrRanges, atac_chr1)

comparison_df <- data.frame(
  Methylation_Diff = dmrRanges$meandiff[queryHits(overlaps)],
  ATAC_logFC = atac_chr1$logFC[subjectHits(overlaps)]
)

ggplot(comparison_df, aes(x = Methylation_Diff, y = ATAC_logFC)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  geom_smooth(method = "lm", color = "red", linetype = "dashed") + # Adds a trendline
  theme_minimal() +
  labs(
    title = "DNA Methylation vs. Chromatin Accessibility (chr1)",
    x = "Change in Methylation (meandiff)",
    y = "Change in Accessibility (ATAC logFC)"
  )
```

To assess the agreement between the two modalities, I identified the genomic regions where the differentially methylated regions (DMRs) physically overlapped with the differential ATAC-seq peaks on chromosome 1. By plotting the change in methylation (meandiff) against the change in accessibility (logFC), it shows that there is a negative correlation (indicated by the trendline). This demonstrates that as DNA methylation increases, chromatin accessibility tends to decrease, which is as expected that methylation is a repressive mark on the DNA.